Small induced universal graphs

### 2021-08-27 paper/paper.tex
Start writing up better bounds for f(6)

\documentclass[12pt]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{amsthm}
\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
%

@@

To give an upper bound of 15 on $f(6)$, Figure~\ref{fig:adjmat15} shows the
adjacency matrix of a 15-vertex graph that is induced universal for the family
of all graphs on six vertices.  This was generated using a specialised random
graph generator.  We number the vertices from zero; the generator makes the
$k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
independent set.  Each possible edge that is not involved in either the clique
or the independent set is added with probability $1/2$.  It was only necessary
to generate 53 graphs with this generator to find a solution.

\begin{figure}[h]
\centering
\verb|0 1 1 1 1 1 0 0 0 1 1 0 0 0 1| \\
\verb|1 0 1 1 1 1 0 0 0 0 1 1 1 1 1| \\
\verb|1 1 0 1 1 1 1 0 0 1 0 0 1 0 0| \\
\verb|1 1 1 0 1 1 1 1 1 1 0 0 0 0 0| \\
\verb|1 1 1 1 0 1 1 1 0 1 1 1 1 0 0| \\
\verb|1 1 1 1 1 0 0 0 0 0 0 0 0 0 1| \\
\verb|0 0 1 1 1 0 0 0 0 0 0 0 0 1 0| \\
\verb|0 0 0 1 1 0 0 0 0 0 0 1 1 1 1| \\
\verb|0 0 0 1 0 0 0 0 0 0 0 1 1 1 0| \\
\verb|1 0 1 1 1 0 0 0 0 0 0 1 0 0 0| \\
\verb|1 1 0 0 1 0 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 0 0 1 0 0 1 1 1 0 0 0 1 0| \\
\verb|0 1 1 0 1 0 0 1 1 0 1 0 0 1 1| \\
\verb|0 1 0 0 0 0 1 1 1 0 0 1 1 0 0| \\
\verb|1 1 0 0 0 1 0 1 0 0 1 0 1 0 0|
\caption{The adjacency matrix of a 15-vertex induced universal graph for the family of all
six-vertex graphs}
\label{fig:adjmat15}
\end{figure}

For $k > 5$, it is straightforward to increase by 1 the lower bound by ``bof'' from the previous
section.

\begin{proposition}
    $f(k) > 2k + 1$ for all $k > 5$.
\end{proposition}
\begin{proof}
    Suppose, for contradition, that $G$ is an induced universal graph on $2k$
    vertices for the family of all graphs on $k$ vertices.  Graph $G$ must have
    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set must
    overlap by exactly one vertex (otherwise, as in the proof by ``bof'' from
    the previous section, $G$ would be a split graph and would not have $C_4$
    as an induced subgraph).   Thus we can partition the vertices of $G$ into
    three sets: a clique $S_1$, an independent set $S_2$, and a set $S_3$
    containing the single remaining vertex.

    Let $S$ be a set of vertices such that the subgraph of $G$ induced by $S$
    is isomorphic to the cycle graph $C_k$.  $S$ may contain at most two vertices
    from $S_1$, and these must be adjacent in the induced cycle.  Trivially,
    $S$ can contain at most one vertex from $S_3$.  Hence, $S$ must contain at
    least $k-3$ vertices from $S_2$.  At least two of these vertices must be adjacent
    in the induced $C_k$.  Since $S_2$ is an independent set in $G$, we have a contradiction.
\end{proof}

Thus, we have that $13 \leq f(6) \leq 15$.

@@

\label{fig:trees}
\caption{A universal graph for the family of all
trees with 6 vertices}
\end{figure}

>>>>

\documentclass[12pt]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{amsthm}
\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}

\usepackage{caption}
\captionsetup{width=0.8\textwidth,font=footnotesize}
%

@@

For $k > 6$, we can increase by 2 the lower bound by ``bof'' from the previous
section.

\begin{proposition}\label{f6proposition}
    $f(k) \geq 2k + 2$ for all $k > 5$.
\end{proposition}
\begin{proof}

    Suppose that $G$ is an induced universal graph
    for the family of all graphs on $k$ vertices, and that $G$ has no
    more than $2k + 1$ vertices.  Graph $G$ must have
    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set may
    overlap by no more than one vertex.
    Therefore it is possible to partition the vertices of $G$ into
    three sets: a clique $S_1$,
    an independent set $S_2$, and a set $S_3$
    containing at most 2 vertices.

    We will show that that $G$ can contain as an induced subgraph at most
    one of the complementary graphs in Figure~\ref{fig:boundproof}, which we
    call $H$ and $H'$ respectively.
    First, let $G_1$ be an induced
    subgraph of $G$ that is isomorphic to $H$.  Since there are
    no edges between the two three-vertex cliques in $G_1$, it must be the case that the
    vertex set of at least one of these cliques does not intersect $S_1$.
    Since $S_2$ is an independent set in $G$, this clique must have exactly
    one vertex in $S_2$ and two vertices in $S_3$.  We can deduce, then, that
    $S_3$ contains exactly two vertices, and that these vertices are adjacent in $G$.

% \begin{tikzpicture}[>=latex',line join=bevel]
%   \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
%   \pgfsetlinewidth{.5bp}
%   \node at (0,2.9) (1) {};
%   \node at (0,2) (2) {};
%   \node at (0,1.1) (3) {};
%   \node at (1.1,2.9) (4) {};
%   \node at (1.1,2) (5) {};
%   \node at (1.1,1.1) (6) {};
%   \draw (1) -- (2) -- (3);
%   \draw (3) to [out=140,in=220] (1);
%   \draw (4) -- (5) -- (6);
%   \draw (6) to [out=40,in=-40] (4);
%   \draw (1) -- (4);
%   \draw (1) -- (5);
%   \draw (1) -- (6);
%   \draw (2) -- (4);
%   \draw (2) -- (5);
%   \draw (2) -- (6);
%   \draw (3) -- (4);
%   \draw (3) -- (5);
%   \draw (3) -- (6);
% \end{tikzpicture}
% \qquad \qquad
% \begin{tikzpicture}
%   \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
%   \pgfsetlinewidth{.5bp}
%   \node at (0,2.9) (1) {};
%   \node at (0,2) (2) {};
%   \node at (0,1.1) (3) {};
%   \node at (1.1,2.9) (4) {};
%   \node at (1.1,2) (5) {};
%   \node at (1.1,1.1) (6) {};
% \end{tikzpicture}
% \qquad \qquad
\begin{tikzpicture}[>=latex',line join=bevel]
  \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
  \pgfsetlinewidth{.5bp}
  \node at (0,2.9) (1) {};
  \node at (0,2) (2) {};
  \node at (0,1.1) (3) {};
  \node at (1.1,2.9) (4) {};
  \node at (1.1,2) (5) {};
  \node at (1.1,1.1) (6) {};
  \draw (1) -- (2) -- (3);
  \draw (3) to [out=140,in=220] (1);
  \draw (4) -- (5) -- (6);
  \draw (6) to [out=140,in=220] (4);
\end{tikzpicture}
\qquad \qquad
\begin{tikzpicture}[>=latex',line join=bevel]
  \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
  \pgfsetlinewidth{.5bp}
  \node at (0,2.9) (1) {};
  \node at (0,2) (2) {};
  \node at (0,1.1) (3) {};
  \node at (1.1,2.9) (4) {};
  \node at (1.1,2) (5) {};
  \node at (1.1,1.1) (6) {};
  \draw (1) -- (4);
  \draw (1) -- (5);
  \draw (1) -- (6);
  \draw (2) -- (4);
  \draw (2) -- (5);
  \draw (2) -- (6);
  \draw (3) -- (4);
  \draw (3) -- (5);
  \draw (3) -- (6);
\end{tikzpicture}
%%
\caption{If $k > 5$ and $G$ is a graph on $2k + 1$ or fewer vertices, then
$G$ cannot have as induced subgraphs all four of the following graphs: the
clique $K_k$, the independent set $I_k$, and
the two graphs shown. See the the proof of Proposition~\ref{f6proposition}}.
\label{fig:boundproof}
\end{figure}

    Now consider graph $H'$.  Since $H'$ is an induced subgraph of $G$, it
    follows by taking complements of $H'$ and $G$ that $H$ is an induced
    subgraph of $G'$ (the complement of $G$).  We can repeat the argument
    of the previous paragraph with the roles of $S_1$ and $S_2$ reversed to
    show that the two vertices in $S_3$ must be adjacent in the complement of
    $G$, and therefore not adjacent in $G$.  Since we previously showed that
    these vertices are adjacent in $G$, we have a contradiction.
\end{proof}

To give a lower bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
adjacency matrix of a 14-vertex graph that is induced universal for the family
of all graphs on six vertices.  This was generated using a specialised random
graph generator.  We number the vertices from zero; the generator makes the
$k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
independent set.  Each possible edge that is not involved in either the clique
or the independent set is added with probability $1/2$.  It was only necessary
to generate 53 graphs with this generator to find a solution TODO fixme.

TODO: give more details of generator

\begin{figure}[h]
\centering
\small
\verb|0 1 1 1 1 1 0 1 1 0 0 0 1 0| \\
\verb|1 0 1 1 1 1 1 1 1 1 1 0 0 1| \\
\verb|1 1 0 1 1 1 0 1 1 0 1 0 1 0| \\
\verb|1 1 1 0 1 1 1 0 0 0 1 1 0 0| \\
\verb|1 1 1 1 0 1 0 1 0 0 0 0 0 0| \\
\verb|1 1 1 1 1 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 0 1 0 0 0 0 0 0 0 1 1 1| \\
\verb|1 1 1 0 1 0 0 0 0 0 0 1 1 0| \\
\verb|1 1 1 0 0 0 0 0 0 0 0 0 1 0| \\
\verb|0 1 0 0 0 0 0 0 0 0 0 1 0 1| \\
\verb|0 1 1 1 0 0 0 0 0 0 0 1 1 0| \\
\verb|0 0 0 1 0 1 1 1 0 1 1 0 0 1| \\
\verb|1 0 1 0 0 0 1 1 1 0 1 0 0 0| \\
\verb|0 1 0 0 0 1 1 0 0 1 0 1 0 0|
\caption{The adjacency matrix of a 14-vertex induced universal graph for the family of all
six-vertex graphs}
\label{fig:adjmat14}
\end{figure}

Thus, we have that $f(6) = 14$.

@@

\caption{A universal graph for the family of all
trees with 6 vertices}
\label{fig:trees}
\end{figure}
